---
title: "Introduction to bridging Olink® NPX datasets"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Introduction to bridging Olink® NPX datasets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)
```

Individual Olink® NPX datasets are normalized using either plate control
normalization or intensity normalization methods. Plate control normalization is
generally used for single plate projects or for Explore HT projects, while
intensity normalization is generally used for multi-plate projects.
Additionally, intensity normalization method assumes that all samples within a
project are fully randomized.

In the case where all samples within a project are not fully randomized, or when
a study is separated into separate batches, an additional normalization step is
needed to allow the data to be comparable, since NPX is a relative measurement.
The joint analysis of two or more Olink® NPX datasets often requires an
additional batch correction step to remove technical variations, which is
referred to as overlapping sample reference normalization, bridge normalization,
or just simply bridging.

Batches are groups of samples run at different times. They may or may not use
the same lot of reagents, however bridging is recommended because good
randomization across these sample batches is unlikely.

Bridging is also needed if Olink® NPX datasets are:

-   plate control normalized only and run conditions (e.g lab and reagent lots)
    have changes.

-   intensity normalized but from two different sample populations.

To bridge two or more Olink® NPX datasets, bridging samples are needed to
calculate the assay-specific adjustment factors between datasets. Bridging
samples are shared samples among datasets - that is that samples that are
analyzed in both datasets. The recommended number of bridging samples are shown
in the table below. Olink® NPX datasets without shared samples should not be
combined using the bridging approach described below.

```{r brnrtab, message = FALSE, echo = FALSE}
dplyr::tibble(
  Platform = c("Target 96",
               paste0("Explore 384: \n",
                      "Cardiometabolic, Inflammation, ",
                      "Neurology, and Oncology"),
               paste0("Explore 384: \n",
                      "Cardiometabolic II, Inflammation II,",
                      "Neurology II, and Oncology II"),
               "Explore HT",
               "Reveal",
               "Explore 3072 to Explore HT",
               "Explore 3072 to Reveal"),
  BridgingSamples = c("8-16",
                      "8-16",
                      "16-24",
                      "16-32",
                      "16-24",
                      "40-64",
                      "32-48")
) |>
  kableExtra::kbl(
    booktabs = TRUE,
    digits = 2L,
    caption = paste("Table 1. Recommended number of bridging samples for Olink",
                    "platforms")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "center",
    latex_options = "HOLD_position"
  )
```

The following tutorial is designed to give you an overview of the kinds of data
combining methods that are possible using the Olink® bridging procedure. Before
starting bridging, it is important to check if the same sample IDs were assigned
to the bridging samples.

## Selecting bridging samples

Prior to running the second study, bridging samples must be selected from the
reference study and be added to the second study. These samples can be selected
using the `olink_bridgeselector()` function in Olink Analyze. The bridge
selection function will select a number of bridge samples based on the reference
data. This function selects samples which passes QC and have high detectability.
In the case of where detectability cannot be calculated from the test data set
(ex: Explore HT data), the function will only select samples which pass QC.
External controls are not selected as bridge samples as they are not necessarily
representative of the study and therefore may not cover the dynamic range of
assays that would be expressed within the samples. Note that due to naming
convention differences, it is necessary to exclude the control samples using
either `SampleType == "SAMPLE"` if available or `stringr::str_detect()` as shown
below.

To select samples across the range of the data, the samples are ordered by mean
NPX value and selected across this range. When running the selector, the
sampleMissingFreq value represents a maximum percentage of data below LOD
allowed per sample. When running plasma on smaller panels, such as Target 96,
`sampleMissingFreq = 0.10` can be a good starting point. Larger panels such as
Explore HT have many proteins that are only expressed in certain diseases or
matrices and therefore more data below LOD is expected. In this case
`sampleMissingFreq = 0.5` can be a good starting point. Note that these values
may need to be adjusted based on the Olink product, sample matrix, and
biological context of the samples, for example, specific disease types.
`sampleMissingFreq = 1` will provide a list of all eligible samples, regardless
of the % of data below LOD and the function output provides the percent of
assays below LOD (`PercAssaysBelowLOD`) which can help inform a good starting
point for a specific study.

In this example we will demonstrate how to select 16 bridging samples using
`npx_data1` which will act as the reference data. The selected bridge samples
are displayed in Table 2.

```{r bridge_sample_selection_example, echo = TRUE, eval = TRUE}
bridge_samples <- OlinkAnalyze::npx_data1 |>
  # Excluding control samples. Naming convention may differ.
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL",
      negate = TRUE
    )
  ) |>
  OlinkAnalyze::olink_bridgeselector(
    sampleMissingFreq = 0.1,
    n = 16L
  )
```

```{r bridge_sample_selection, echo = FALSE}
OlinkAnalyze::npx_data1 |>
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL",
      negate = TRUE
    )
  ) |>
  OlinkAnalyze::olink_bridgeselector(
    sampleMissingFreq = 0.1,
    n = 16L
  ) |>
  kableExtra::kbl(
    booktabs = TRUE,
    digits = 2L,
    caption = "Table 2. Selected Bridging Samples"
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "center",
    latex_options = "HOLD_position"
  )
```

It's important to make sure that the select bridge samples are representative of
the overall samples within the study. This can be done by generating a PCA plot
with the following code and make sure that the bridge samples are evenly
dispersed among the other samples, as shown in Figure 1.

```{r captions, echo = FALSE}
f1 <- paste("Figure 1. PCA plot of bridging samples and other samples in",
            "npx_data1. Control samples are excluded from the PCA plot.")

f2 <- paste("Figure 2. Density plot of NPX distribution in both datasets",
            "before bridging.")

f3 <- paste("Figure 3. PCA plot of both datasets before bridging.")

f4 <- paste("Figure 4. Density plot of NPX distribution in both datasets after",
            "bridging.")

f5 <- paste("Figure 5. Histogram of adjustment factors in normalized data from",
            "Project `data2`.")

f6 <- paste("Figure 6. Violin plot of CHL1 in both datasets prior to bridging.",
            "Bridge samples are indicated by black points.")

f7 <- paste("Figure 7. Density plot of inter-project CV before and after",
            "bridging.")

f8 <- paste("Figure 8. PCA plot of both datasets after bridging.")
```

```{r bridge_sample_selection_example_pca, echo = TRUE, eval = TRUE, fig.cap = f1}
OlinkAnalyze::npx_data1 |>
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONT",
      negate = TRUE
    )
  ) |>
  dplyr::mutate(
    Bridge = dplyr::if_else(
      .data[["SampleID"]] %in% bridge_samples$SampleID,
      "Bridge",
      "Sample"
    )
  ) |>
  OlinkAnalyze::olink_pca_plot(
    color_g = "Bridge"
  )
```

## Setup bridging datasets

Bridging datasets are standard Olink® NPX tables. They can be loaded using
`read_NPX()` function with default Olink Software NPX file as input.

```{r, message = FALSE, eval = FALSE, echo = TRUE}
data1 <- OlinkAnalyze::read_NPX(
  filename = "~/NPX_file1_location.xlsx"
)
data2 <- OlinkAnalyze::read_NPX(
  filename = "~/NPX_file2_location.xlsx"
)
```

To demonstrate how bridging works, we will use the example datasets (`npx_data1`
and `npx_data2`) from **Olink Analyze** package. This workflow also uses
functions from the `dplyr`, `stringr`, and `ggplot2` packages.

## Check bridging datasets

First, confirm that there are overlapping sample IDs within the study. It is
important that the sample IDs are the same in both NPX files as shown in Table
3. Note that while external controls may share the same SampleID in both
datasets, theses samples should not be treated as bridging samples. External
control samples often share the same naming convention across datasets but may
represent different samples due to reagent batch differences. Additionally,
external control samples are unlikely to cover the dynamic range of assays
expressed in the study samples.

```{r, eval = FALSE}
dplyr::tibble(
  SampleID = intersect(
    x = OlinkAnalyze::npx_data1$SampleID,
    y = OlinkAnalyze::npx_data2$SampleID
  )
) |>
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL_SAMPLE",
      negate = TRUE
    )
  )
```

```{r, echo = FALSE}
dplyr::tibble(
  SampleID = intersect(
    x = OlinkAnalyze::npx_data1$SampleID,
    y = OlinkAnalyze::npx_data2$SampleID
  )
) |>
  # Remove control samples
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL_SAMPLE",
      negate = TRUE
    )
  ) |>
  kableExtra::kbl(
    booktabs = TRUE,
    digits = 2L,
    caption = "Table 3. Overlapping Bridge Samples"
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "center",
    latex_options = "HOLD_position"
  )
```

Then, gain an overview of the datasets that are going to be bridged. For
example, plot and compare NPX distribution between datasets, as shown in Figure
2. By having a sense of how the studies compared to each other before bridging,
we can then determine the success of the bridging process post bridging. Figure
2 shows a large overlap between data1 and data2, however there are still some
differences as indicated by the blue and red areas on the edge of the
distribution.

```{r check, message = FALSE, fig.cap = f2}
# Load datasets
npx_1 <- OlinkAnalyze::npx_data1 |>
  dplyr::mutate(
    Project = "data1"
  )
npx_2 <- OlinkAnalyze::npx_data2 |>
  dplyr::mutate(
    Project = "data2"
  )

npx_df <- dplyr::bind_rows(npx_1, npx_2)

# Plot NPX density before bridging normalization
npx_df |>
  dplyr::mutate(
    Panel = gsub(
      pattern = "Olink ",
      replacement = "",
      x = .data[["Panel"]]
    )
  ) |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = .data[["NPX"]],
      fill = .data[["Project"]]
    )
  ) +
  ggplot2::geom_density(
    alpha = 0.4
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(.data[["Panel"]])
  ) +
  OlinkAnalyze::olink_fill_discrete(
    coloroption = c("red", "darkblue")
  ) +
  OlinkAnalyze::set_plot_theme() +
  ggplot2::ggtitle(
    label = "Before bridging normalization: NPX distribution"
  ) +
  ggplot2::theme(
    axis.title.x = ggplot2::element_blank(),
    axis.title.y = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(size = 16L),
    legend.title = ggplot2::element_blank(),
    legend.position = "top"
  )
```

Use a PCA plot as shown in Figure 3 to visualize sample-to-sample distance
before bridging. Typically the project dataset accounts for most of the observed
variation within the combined datasets at this point. Control samples are
removed from the PCA plot because they are not expected to be representative of
the rest of the dataset, and therefore the variation between sample and control
samples could skew the PCA, making it more challenging to see differences
between projects. Additional, the PCA function does not support multiple samples
with the same SampleID, which is also why the bridging samples in the second
project are appended with "\_new". In Figure 3, data1 and data2 separate along
PC1, indicating batch differences between the two datasets.

```{r pca1, message = FALSE, fig.cap = f3}
#### Extract bridging samples

overlapping_samples <- dplyr::tibble(
  SampleID = intersect(
    x = npx_1$SampleID,
    y = npx_2$SampleID
    )
  ) |>
  # Remove control samples
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL_SAMPLE",
      negate = TRUE
    )
  ) |>
  dplyr::pull(
    .data[["SampleID"]]
  )

npx_before_br <- OlinkAnalyze::npx_data1 |>
  # Remove control samples
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL_SAMPLE",
      negate = TRUE
    )
  ) |>
  dplyr::mutate(
    Type = dplyr::if_else(
      .data[["SampleID"]] %in% .env[["overlapping_samples"]],
      "data1 Bridge",
      "data1 Sample"
    )
  ) |>
  dplyr::bind_rows(
    OlinkAnalyze::npx_data2 |>
      # Remove control samples
      dplyr::filter(
        stringr::str_detect(
          string = .data[["SampleID"]],
          pattern = "CONTROL_SAMPLE",
          negate = TRUE
        )
      ) |>
      dplyr::mutate(
        Type = dplyr::if_else(
          .data[["SampleID"]] %in% .env[["overlapping_samples"]],
          "data2 Bridge",
          "data2 Sample"
        )
      ) |>
      dplyr::mutate(
        SampleID = dplyr::if_else(
          .data[["SampleID"]] %in% .env[["overlapping_samples"]],
          paste0(.data[["SampleID"]], "_new"),
          .data[["SampleID"]]
        )
      )
  )

if (requireNamespace(package = "ggpubr", quietly = TRUE)) {
  ### PCA plot
  OlinkAnalyze::olink_pca_plot(
    df = npx_before_br,
    color_g = "Type",
    byPanel = TRUE
  )
}
```

## Perform bridging between two data sets

We can use `olink_normalization_bridge()` function to bridge two datasets. The
bridging procedure is to first calculate **the median of the paired NPX**
**differences** per assay between the bridging samples as adjustment factor then
use these adjustment factors to adjust NPX values between two datasets. In this
process, one dataset is considered the reference dataset (`df1`) and its NPX
values remain unaltered. The other dataset is considered the new dataset (`df2`)
and is adjusted to the reference dataset based on the adjustment factors.

The output from `olink_normalization_bridge()` function is a NPX table with
adjusted NPX value in the column `NPX`, as shown in Table 4.
`olink_normalization_bridge()` is a wrapper for and supersedes
`olink_normalization()` .

`olink_normalization_bridge()` creates a new column `Project` to distinguish
between reference dataset from the other dataset. It is up to the user to define
which dataset is the reference dataset and specify the names of the bridge
samples. The resulting dataset will contain the reference dataset, which will be
identical to the input reference data, with adjustment factors of 0, and the
newly bridged dataset.

NOTE : The alternative quantification columns (for example: Quantified value,
count, and Ct data) are not normalized.

```{r bridging, message = FALSE}
# Find shared samples
npx_1 <- OlinkAnalyze::npx_data1 |>
  dplyr::mutate(
    Project = "data1"
  )
npx_2 <- OlinkAnalyze::npx_data2 |>
  dplyr::mutate(
    Project = "data2"
  )

overlap_samples <- dplyr::tibble(
  SampleID = intersect(
    x = npx_1$SampleID,
    y = npx_2$SampleID
  )
) |>
  # Remove control samples
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL_SAMPLE",
      negate = TRUE
    )
  ) |>
  dplyr::pull(
    .data[["SampleID"]]
  )

overlap_samples_list <- list(
  "DF1" = overlap_samples,
  "DF2" = overlap_samples
)

# Perform Bridging normalization
npx_br_data <- OlinkAnalyze::olink_normalization_bridge(
  project_1_df = npx_1,
  project_2_df = npx_2,
  bridge_samples = overlap_samples_list,
  project_1_name = "data1",
  project_2_name = "data2",
  project_ref_name = "data1"
)
```

```{r norm_data_table, echo = FALSE}
npx_br_data |>
  dplyr::slice_head(
    n = 10L
  ) |>
  kableExtra::kbl(
    booktabs = TRUE,
    digits = 1L,
    caption = "Table 4. First 10 rows of combined datasets after bridging."
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    font_size = 10L,
    position = "center",
    latex_options = "HOLD_position"
  ) |>
  kableExtra::scroll_box(
    width = "100%"
  )
```

## Perform bridging with non-matching sample names

`olink_normalization_bridge()` also supports data where the shared samples
(bridge samples) are not named the same in both projects. In this case the
`overlap_sample_list` will contain 2 arrays of equal length where the index of
each entry corresponds to the same sample. For example, if a sample had the
SampleID of `Sample_1_Aliquot_1` in the first batch and `Sample_1_Aliquot_2` in
the second batch, then the overlap sample list should look as the following.

```{r, echo = TRUE, eval = TRUE}
overlap_sample_list <- list(
  "DF1" = c("A1", "A2", "A3", "Sample_1_Aliquot_1"),
  "DF2" = c("A1", "A2", "A3", "Sample_1_Aliquot_2")
)
```

## Perform evaluation of bridging normalization

First, check NPX distribution in datasets after bridging normalization, as shown
in Figure 4. Figure 4 shows a larger overlap in projects than seen in Figure 2,
as indicated by fewer areas of red and blue along the outside of the
distribution.

```{r densitybr, message = FALSE, fig.cap = f4}
# Plot NPX density after bridging normalization

npx_br_data |>
  dplyr::mutate(
    Panel = gsub(
      pattern = "Olink ",
      replacement = "", x = .data[["Panel"]]
    )
  ) |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = .data[["NPX"]],
      fill = .data[["Project"]]
    )
  ) +
  ggplot2::geom_density(
    alpha = 0.4
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(.data[["Panel"]])
  ) +
  OlinkAnalyze::olink_fill_discrete(
    coloroption = c("red", "darkblue")
  ) +
  OlinkAnalyze::set_plot_theme() +
  ggplot2::ggtitle(
    label = "After bridging normalization: NPX distribution"
  ) +
  ggplot2::theme(
    axis.title.x = ggplot2::element_blank(),
    axis.title.y = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(size = 16L),
    legend.title = ggplot2::element_blank(),
    legend.position = "top"
  )
```

Then, summarize number of assays that have adjustment factors in certain ranges.
High adjustment factors can result from variations between projects, such as
panel versions or technical modifications. The cutoff of a deviating adjustment
factor is subjective and depends on a variety of factors including the
distribution of adjustment factors, as shown in Figure 5. While there are a few
assays with adjustment factors between 2 and 4 and -2 and -4, the majority of
the adjustment factors shown in Figure 5 occur between -2 and 2.

Such assays can be visualized individually with violin plots, as shown in Figure
6, and may warrant further investigation to confirm they are still comparable
between projects. For example, if a violin plot exhibits a different range or
truncated distribution, this may suggest that the assay is below LOD or at hook
in one of the data sets. However, as long as the bridge samples are not at hook
or below LOD, this should not impact the bridging quality. For projects with
differing clinical phenotypes, it is more informative to look at the
similarities between the bridging samples than the similarity between the
datasets, as indicated by the overlaid black dots in Figure 6.

```{r dist_adj_fct, message = FALSE, echo = FALSE, fig.cap = f5}
npx_br_data |>
  # Only looking at Project 2 since project 1 is unadjusted
  dplyr::filter(
    .data[["Project"]] == "data2"
  ) |>
  dplyr::select(
    dplyr::all_of(
      c("OlinkID", "Adj_factor")
    )
  ) |>
  dplyr::distinct() |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = .data[["Adj_factor"]]
    )
  ) +
  ggplot2::geom_histogram() +
  OlinkAnalyze::set_plot_theme()
```

The distribution of CHL1 between projects is visualized for demonstration
purposes.

```{r violin_plot, fig.cap = f6}
# Bridge sample data
bridge_samples <- npx_1 |>
  dplyr::bind_rows(
    npx_2
  ) |>
  dplyr::filter(
    .data[["SampleID"]] %in% .env[["overlapping_samples"]]
  ) |>
  dplyr::filter(
    .data[["Assay"]] == "CHL1"
  ) |>
  dplyr::mutate(
    Assay_OID = paste(.data[["Assay"]], .data[["OlinkID"]], sep = "\n")
  )

# Generate violin plot for CHL1
OlinkAnalyze::npx_data1 |>
  dplyr::mutate(
    Project = "data1"
  ) |>
  dplyr::bind_rows(
    OlinkAnalyze::npx_data2 |>
      dplyr::mutate(
        Project = "data2"
      )
  ) |>
  dplyr::filter(
    .data[["Assay"]] == "CHL1"
  ) |>
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL*.",
      negate = TRUE
    )
  ) |>
  dplyr::mutate(
    Assay_OID = paste(.data[["Assay"]], .data[["OlinkID"]], sep = "\n")
  ) |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = .data[["Project"]],
      y = .data[["NPX"]]
    )
  ) +
  ggplot2::geom_violin(
    mapping = ggplot2::aes(
      fill = .data[["Project"]]
    )
  ) +
  ggplot2::geom_point(
    data = bridge_samples,
    position = ggplot2::position_jitter(
      width = 0.1
    )
  ) +
  ggplot2::theme(
    legend.position = "none"
  ) +
  OlinkAnalyze::set_plot_theme() +
  ggplot2::facet_wrap(
    facets = . ~ Assay_OID,
    scales = "free_y"
  )
```

Another way to determine if bridging decreased variability between projects is
to calculate the CV of the control samples across both projects before and after
bridging, as shown in Figure 7. The CV after normalization is expected to be
smaller than the CV prior to normalization, as shown in Figure 7.

Note that the CV calculation formula differs for Olink Target 96 and Olink
Explore products. More information of CV calculation can be found in the
[Olink FAQ](https://olink.com/knowledge/faq).

```{r CV_calculation, fig.cap = f7}
explore_cv <- function(npx, na_rm = FALSE) {
  sqrt(exp((log(2) * sd(npx, na.rm = na_rm)) ^ 2) - 1) * 100
}

t96_cv <- function(npx, na_rm = TRUE) {
  100 * sd(2 ^ npx) / mean(2 ^ npx)
}

tech <- "Target"

cv_before <- npx_1 |>
  dplyr::bind_rows(
    npx_2
  ) |>
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL*."
    )
  ) |>
  dplyr::filter(
    .data[["NPX"]] > .data[["LOD"]]
  ) |>
  dplyr::group_by(
    .data[["OlinkID"]]
  ) |>
  dplyr::mutate(
    CV = dplyr::if_else(
      .env[["tech"]] == "Explore",
      explore_cv(npx = .data[["NPX"]]),
      t96_cv(npx = .data[["NPX"]])
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::distinct(
    .data[["OlinkID"]], .data[["CV"]]
  )

cv_after <- npx_br_data |>
  dplyr::filter(
    stringr::str_detect(
      string = .data[["SampleID"]],
      pattern = "CONTROL*."
    )
  ) |>
  dplyr::filter(
    .data[["NPX"]] > .data[["LOD"]]
  ) |>
  dplyr::group_by(
    .data[["OlinkID"]]
  ) |>
  dplyr::mutate(
    CV = dplyr::if_else(
      .env[["tech"]] == "Explore",
      explore_cv(npx = .data[["NPX"]]),
      t96_cv(npx = .data[["NPX"]])
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::distinct(
    .data[["OlinkID"]], .data[["CV"]]
  )

cv_before |>
  dplyr::mutate(
    Analysis = "Before"
  ) |>
  dplyr::bind_rows(
    cv_after |>
       dplyr::mutate(
         Analysis = "After"
        )
  ) |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = .data[["CV"]],
      fill = .data[["Analysis"]]
      )
    ) +
  ggplot2::geom_density(
    alpha = 0.7
    ) +
  OlinkAnalyze::set_plot_theme() +
  OlinkAnalyze::olink_fill_discrete() +
  ggplot2::theme(
    text = ggplot2::element_text(size = 20L)
  ) +
  ggplot2::xlim(-50L, 400L)
```

Finally, use PCA plot to check whether bridging normalization has effect in
correcting batch effects (Figure 8). In the example below, it is clear that
before bridging samples from data 1 and 2 are divided into separate clusters due
to the batch effects (Figure 3), but after bridging they are shown as one
cluster in the PCA plot (Figure 8). Bridging normalization has sufficiently
removed the batch effects between two data sets.

```{r pca2, message = FALSE, fig.cap = f8}
## After bridging

### Generate unique SampleIDs

npx_after_br <- npx_br_data |>
  dplyr::mutate(
    Type = dplyr::if_else(
      .data[["SampleID"]] %in% .env[["overlapping_samples"]],
      paste(.data[["Project"]], "Bridge"),
      paste(.data[["Project"]], "Sample")
    )
  ) |>
  dplyr::mutate(
    SampleID = paste0(.data[["Project"]],
                      .data[["PlateID"]],
                      .data[["SampleID"]])
  )

if (requireNamespace(package = "ggpubr", quietly = TRUE)) {
  ### PCA plot
  OlinkAnalyze::olink_pca_plot(
    df = npx_after_br,
    color_g = "Type",
    byPanel = TRUE
  )
}
```

## Export Bridged data

Normalized data can be exported using write.table to export long format data.
Note that 2 columns are added during the bridging process, so to have the input
format match the export format the Project and Adj_factor columns will need to
be removed. To export the new project, `dplyr::filter` can be used to filter by
Project.

```{r, eval = FALSE}
new_normalized_data <- npx_br_data |>
  dplyr::filter(
    .data[["Project"]] == "data2"
  ) |>
  dplyr::select(
    -dplyr::all_of(
      c("Project", "Adj_factor")
    )
  ) |>
  write.table(
    file = "New_Normalized_NPX_data.csv",
    sep = ";"
  )
```

## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© `r format(Sys.Date(), "%Y")` Olink Proteomics AB, part of Thermo Fisher
Scientific.

Olink products and services are For Research Use Only. Not for use in diagnostic
procedures.

All information in this document is subject to change without notice. This
document is not intended to convey any warranties, representations and/or
recommendations of any kind, unless such warranties, representations and/or
recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions based on
this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB. All
third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and patent
applications [https://www.olink.com/patents/](https://olink.com/patents/).
